R markdown file for PeerAssessment 1 of Reproducible Research
==============================================================
This R markdown file will read in data in a file called "activity.csv"
There are three columns in the data file :steps, date and interval.


## Setting the global echo option to TRUE
======================================
``` {r setoptions,echo=TRUE}
    opts_chunk$set(echo= TRUE, results ="asis")
```

## Loading the data
=================

```{r}
RawData<-read.csv("activity.csv")
```

## Processing the data
===================
```{r}

PData<-RawData
#Determining the total number of rows in original data set
NROW<-nrow(PData)
# Dealing with the dates
PData$date<-as.character(PData$date)
PData$date<-as.Date(PData$date,"%Y-%m-%d")
#Getting the unique dates
UDates<-unique(PData$date)
#Getting the number of days
NODs<-length(UDates)
# Converting to the corresponding day
PData$Day<-weekdays(PData$date)
# Steps and Interval
PData$steps<-as.numeric(PData$steps)
PData$interval<-as.numeric(PData$interval)
# Removing the NAs
PMData<-na.omit(PData)



```

## Plots corresponding to the loaded data
============================================

### 1.Histogram of total Number of Steps taken in a day & Numerical values of Mean and Median
=============================================================================================
Histogram is plotted by omitting the missing values  
Mean and Median is calculated by omitting the missing values


```{r}
  #Plotting the histogram of the total number of steps taken each day
  TotalHist<-tapply(PMData$steps,PMData$date,sum)
  hist(TotalHist,main="Histogram with NAs excluded",xlab='Total number of steps in a day')
  #Mean of the steps excluding the NAs
  MeanStepsWONA <-mean(na.omit(PData$steps))
  MeanStepsWONA  
  #Median of the steps excluding the NAs
  MedianStepsWONA<-median(na.omit(PData$steps))
  MedianStepsWONA

```

### 2.Average daily activity pattern
=============================================
```{r}
  UIDs<-unique(PData$interval)
  LengthUIDS<-length(UIDs)
  AvgDate<-tapply(PMData$steps,PMData$interval,mean)
  RefMatrix<-data.frame(UIDs,AvgDate)
  RefMatrix$AvgDate<-as.numeric(RefMatrix$AvgDate)
  plot(UIDs,AvgDate,type="l",main="Average daily activity pattern",xlab="Interval Number",
       ylab="Number of steps")
  #The five minute interval having the maximum value for the average number of steps
  MaxAvgIndex<-which.max(as.numeric(AvgDate))
  MaxAvgInterval<-UIDs[MaxAvgIndex]
  #5 minute interval having the max average number of steps
  MaxAvgInterval
  
```
  

### 3.Inputting missing values
=============================================
a.Total number of missing values in the dataset  
```{r}
NumberNA<-sum(is.na(PData))
#Number of missing values
NumberNA   
```  

b.Filling in the missing values and creating a new Data Set  s
  
```{r}
#Missing value is replaced by the average of the corresponding 5 minute interval
PM2Data<-PData
for (i in 1:NROW){
  if(is.na(PM2Data$steps[i]))
    {
            PM2Data$steps[i]<-AvgDate[match(PM2Data$interval[i],UIDs)]
    }
}

```
c.Histogram, mean and median of new data set  
```{r}
  NewTotalHist<-tapply(PM2Data$steps,PM2Data$date,sum)
  hist(NewTotalHist,main="Histogram with replacements for NAs ",xlab='Total number of steps in a day')
  #Mean of the steps excluding the NAs
  NewMeanSteps<-mean(PM2Data$steps)
  NewMeanSteps  
  #Median of the steps excluding the NAs
  NewMedianSteps<-median(PM2Data$steps)
  NewMedianSteps
```

### 4.Difference in activity patterns on wwekdays and weekends
===========================================================

```{r}
#Coding an extra column indicating a weekday or weekend
  PM2Data<-within(PM2Data,{
    daycat<-NA
    daycat[PM2Data$Day=="Monday"]<-"Weekday"
    daycat[PM2Data$Day=="Tuesday"]<-"Weekday"
    daycat[PM2Data$Day=="Wednesday"]<-"Weekday"
    daycat[PM2Data$Day=="Thursday"]<-"Weekday"
    daycat[PM2Data$Day=="Friday"]<-"Weekday"
    daycat[PM2Data$Day=="Saturday"]<-"Weekend"
    daycat[PM2Data$Day=="Sunday"]<-"Weekend"
  }
    )
#Subsetting based on weekday or weekend
DataWeekday<-subset(PM2Data,PM2Data$daycat=="Weekday")
DataWeekend<-subset(PM2Data,PM2Data$daycat=="Weekend")
#Finding the average for each time interval.ie., averaged across the days
AvgWeekday<-tapply(DataWeekday$steps,DataWeekday$interval,mean)
AvgWeekend<-tapply(DataWeekend$steps,DataWeekend$interval,mean)
AvgWeekdayDF<-data.frame(UIDs,AvgWeekday)
AvgWeekendDF<-data.frame(UIDs,AvgWeekend)
#Plotting of the data based on weekday or weekend
par(mfrow = c(2,1))
par(mar=c(4,4,1,1))
plot(AvgWeekdayDF$UIDs,AvgWeekdayDF$AvgWeekday,type="l",
     main="Weekday",xlab="",ylab="Number of Steps")
plot(AvgWeekendDF$UIDs,AvgWeekendDF$AvgWeekend,type="l",
     main="Weekend",xlab="",ylab="Number of Steps")

```

